---
title: Explore Economic Portrait of Greater Philadelphia
format: 
  html: 
    toc: true
    toc-depth: 2
    code-fold: true
    code-summary: "Show code" 
    theme: flatly
author: 
  - name: Yaelle Tang
    email: yueyang_tang@berkeley.edu
    url: https://yaelle-tang.github.io/personal-website/
date: last-modified
date-format: "YYYY-MM-DD HH:mm"
execute: 
  warning: false # suppress warnings
---

## Potential sources

* [The Inquirer - Delawrae County](https://www.inquirer.com/topic/delaware-county/)

1. [Ryotaro Tashiro](https://www.philadelphiafed.org/our-people/ryotaro-tashiro), outreach economist at the Federal Reserve Bank of Philadelphia, who tracks economic trends across the greater Philladelphia area. He is responsible for the [Chamber of Commerce for Greater Philadelphia Economic Outlook Survey](https://www.philadelphiafed.org/surveys-and-data/regional-economic-analysis/chamber-of-commerce-for-greater-philadelphia-economic-outlook-survey). He could provide context on the broader manufacturing decline across southeastern Pennsylvania. 
  - [Worker wages are growing faster in the Philadelphia region than any other metro area - May 11, 2023](https://www.inquirer.com/jobs/labor/unemployment-wages-philadelphia-region-labor-statistics-20230511.html)
2. [Mike Boyer](https://www.dvrpc.org/about/staff/) (215) 238-2848, Director of Regional Planning, Delaware Valley Regional Planning Commission
 - [WHERE HAVE ALL THE WAGES GONE? Six years postrecession, incomes are still down across the region, and in some areas by a lot. December 6, 2015](https://doccenter.inquirer.com/document/10708278)
 

```{r}
#| label: set-up
#| include: false

# install.packages("bea.R")

# Load libraries
library(tidyverse)
library(readxl)
library(janitor)
library(rio)
library(bea.R)
library(jsonlite)
library(dplyr)

```

```{r}
#| label: API-key
#| include: false


beakey <- "94D82C6D-AE74-41C9-ACC2-C2896A580F8E"
```

```{r}
#| label: check the description of all the datasets
#| include: false


url_1 <- paste0(
  "https://apps.bea.gov/api/data/?",
  "UserID=", beakey,
  "&method=GetParameterValues",
  "&DataSetName=Regional",
  "&ParameterName=TableName"
)

description <- fromJSON(url_1)$BEAAPI$Results$ParamValue

```

# Data Memo

## Retrieve GDP Data

::: {.callout-note collapse="false"}
## About the data
GDP CAGDP9 - Real GDP by industry in chained dollars [Source Page](https://apps.bea.gov/itable/?ReqID=70&step=1&_gl=1*17c6s0v*_ga*ODg5OTIwODkyLjE3ODExOTg3ODc.*_ga_J4698JNNFT*czE3ODEyMDE1OTEkbzIkZzEkdDE3ODEyMDE2MzYkajE1JGwwJGgw) 

And more detail: [Regional GDP & Personal Income](https://www.bea.gov/itable/regional-gdp-and-personal-income?_gl=1*1sual4s*_ga*ODg5OTIwODkyLjE3ODExOTg3ODc.*_ga_J4698JNNFT*czE3ODEyMDE1OTEkbzIkZzEkdDE3ODEyMDE2OTEkajU4JGwwJGgw)
:::

```{r}
#| label: Retrieve CAGDP9

beaSpec <- list(
  "UserID"      = beakey,
  "Method"      = "GetData",
  "DataSetName" = "Regional",
  "TableName"   = "CAGDP9",
  "LineCode"    = "1",
  "GeoFips"     = "42045",
  "Year"        = "ALL"
)

df <- beaGet(beaSpec, asWide = FALSE)
```

```{r}

#| label: Test - Bucks county


beaSpec <- list(
  "UserID"      = beakey,
  "Method"      = "GetData",
  "DataSetName" = "Regional",
  "TableName"   = "CAGDP9",
  "LineCode"    = "ALL",
  "GeoFips"     = "42017",
  "Year"        = "ALL"
)


# "GeoFips"     = "42017,42091,42029,42045,42101,34005,34007,34015"
gdp_bucks <- beaGet(beaSpec, asWide = FALSE)

gdp_bucks
```


```{r}
#| label: Get all the counties GDP by industry in all years

fips_list <- c("42017", "42091", "42029", "42045", "42101", "34005", "34007", "34015")

df_all <- lapply(fips_list, function(fips) {
  Sys.sleep(1)
  beaSpec <- list(
    "UserID"      = beakey,
    "Method"      = "GetData",
    "DataSetName" = "Regional",
    "TableName"   = "CAGDP9",
    "LineCode"    = "ALL",
    "GeoFips"     = fips,
    "Year"        = "ALL"
  )
  beaGet(beaSpec, asWide = FALSE)
}) |> 
  bind_rows() |>
  clean_names() |>
  mutate(
    data_value = as.numeric(data_value),
    time_period = as.integer(time_period),
    unit_mult = as.integer(unit_mult))
```

```{r}
#| label:  Clean dataet


df_clean <- df_all |>
  mutate(line_code = str_extract(code, "(?<=-)\\d+")) |>  # AI-assisted
  select(geo_fips, geo_name, time_period, line_code, description, cl_unit, data_value)

glimpse(df_clean)
```
## GDP changes 2001-2025 (Real GDP, thousands of chained 2017 dollars)

```{r}
#| label: GDP changes


df_industry_total <- df_clean |>
  filter(description == "All industry total ") 

# AI-assisted
ggplot(df_industry_total, aes(x = time_period, y = data_value, color = geo_name)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Real GDP by County(2001-2024), Greater Philadelphia",
    x = "Year",
    y = "Real GDP (thousands of chained 2017 dollars)",
    color = "County"
  )

# export data
df_industry_total_wide <- df_industry_total |>
  select(geo_name, time_period, data_value) |>
  pivot_wider(names_from = geo_name, values_from = data_value)

write_csv(df_industry_total_wide, "../data/processed/greater_philly_real_gdp_2001_2024.csv")
```

<iframe title="Real GDP by County (2001-2024), Greater Philadelphia" aria-label="Line chart" id="datawrapper-chart-j1Cv6" src="https://datawrapper.dwcdn.net/j1Cv6/1/" scrolling="no" frameborder="0" style="border: none;" width="600" height="642" data-external="1"></iframe>

```{r}
#| label: Select industry sectors, avoid double counting


df_industry_structure <- df_clean |>
  filter(time_period == 2024) |>
  left_join(
    df_clean |>
      filter(time_period == 2024, line_code == "1") |>
      select(geo_fips, total_gdp = data_value),
    by = "geo_fips"
  ) |> 
  mutate(share = data_value / total_gdp) |>
  filter(line_code != "1")

options(scipen = 999)

# filter

df_industry_filter <- df_industry_structure |>
  filter(line_code %in% c(3,6,10,11,12,34,35,36,45,51,56,60,64,65,69,70,76,79,82,83)) |>
  select(geo_name,line_code,description,data_value,share)

glimpse(df_industry_filter)

# check
df_industry_filter |>
  group_by(geo_name) |>
  summarise(total_share = sum(share, na.rm = TRUE))

```

## What is the industry structure in different counties of Greater Philadelphia in 2024?
```{r} 
#| label: Industry Structure by county in 2024

# Tag "other" industry
df_industry_filter <- df_industry_filter |>
  mutate(
    description_grouped = if_else(share >= 0.05, description, "Other")
  )

# df_industry_filter |> 
#   select(geo_name, data_value, share, description_grouped) |>
#   write_csv("../data/processed/greater_philly_industry_2024.csv")

# AI-assisted
df_industry_filter_wide <- df_industry_filter |>
  select(geo_name, description_grouped, share) |>
  group_by(geo_name, description_grouped) |>
  summarise(share = sum(share), .groups = "drop") |>
  pivot_wider(names_from = description_grouped, values_from = share)

write_csv(df_industry_filter_wide, "../data/processed/greater_philly_industry_2024.csv")
  
```

<iframe title="Industry Structure by County, Greater Philadelphia (2024)" aria-label="Stacked Bars" id="datawrapper-chart-H1mf0" src="https://datawrapper.dwcdn.net/H1mf0/3/" scrolling="no" frameborder="0" style="border: none;" width="700" height="593" data-external="1"></iframe>

## How does the industry GDP contribution change in Delaware County 2001-2024
```{r}
#| label: Industry composition

# filter out Delaware county 
df_delaware <- df_clean |>
  filter(geo_name == "Delaware") |>
  left_join(
    df_clean |>
      filter(geo_name == "Delaware", line_code == "1") |>
      select(time_period, total_gdp = data_value),
    by = "time_period"
  ) |>
  mutate(share = data_value / total_gdp) |>
  filter(line_code %in% c(3,6,10,11,12,34,35,36,45,51,56,60,64,65,69,70,76,79,82,83))|>
  select(geo_name,line_code,time_period,description, share)

# AI-assisted
# calculate average share during 2001-2024, tag "other" label
industry_avg <- df_delaware |>
  group_by(line_code, description) |>
  summarise(avg_share = mean(share), .groups = "drop")

# share >= 0.05
major_industries <- industry_avg |>
  filter(avg_share >= 0.05) |>
  pull(description)

df_delaware <- df_delaware |>
  mutate(description_grouped = if_else(description %in% major_industries, description, "Other"))

df_delaware_wide <- df_delaware |>
  group_by(time_period, description_grouped) |>
  summarise(share = sum(share), .groups = "drop") |>
  pivot_wider(names_from = description_grouped, values_from = share)

write_csv(df_delaware_wide, "../data/processed/delaware_industry_2001-2024.csv")
```

<iframe title="Industry Composition of Delaware County's GDP (2001-2024)" aria-label="Area Chart" id="datawrapper-chart-9S1ZA" src="https://datawrapper.dwcdn.net/9S1ZA/1/" scrolling="no" frameborder="0" style="border: none;" width="650" height="624" data-external="1"></iframe>

## LODES - Commute data

<iframe title="Delaware County Residents Work Places Changes, 2013-2023" aria-label="Symbol map" id="datawrapper-chart-cc2yA" src="https://datawrapper.dwcdn.net/cc2yA/1/" scrolling="no" frameborder="0" style="border: none;" width="640" height="529" data-external="1"></iframe>


::: {.callout-note collapse="false"}
## About the data
- LEHD([Longitudinal Employer-Household Dynamics](https://lehd.ces.census.gov/)): exploring the interactions between workers and firms
- LODES(LEHD Origin-Destination Employment Statistics [Index of /data/lodes](https://lehd.ces.census.gov/data/lodes/) )
- LODES Introduction pdf: https://lehd.ces.census.gov/data/lodes/LODES8/LODESTechDoc8.1.pdf
:::

<iframe title="Delaware County Residents' Work Destinations (absolute number), 2013 vs. 2023" aria-label="Grouped Bars" id="datawrapper-chart-XDBB0" src="https://datawrapper.dwcdn.net/XDBB0/1/" scrolling="no" frameborder="0" style="border: none;" width="650" height="785" data-external="1"></iframe>

<iframe title="Delaware County Residents' Work Destinations (share change), 2013 vs. 2023" aria-label="Arrow Plot" id="datawrapper-chart-KG4Cv" src="https://datawrapper.dwcdn.net/KG4Cv/2/" scrolling="no" frameborder="0" style="border: none;" width="600" height="504" data-external="1"></iframe>



```{r}
#| label: people who both live and work in Delaware County


lodes <- read_csv("../data/raw/lodes_2013-2023.csv")

glimpse(lodes)

lodes <- lodes |>
  mutate(
    w_county = str_sub(w_geocode, 1, 5),
    h_county = str_sub(h_geocode, 1, 5)
  )

lodes_county <- lodes |>
  group_by(year) |>
  summarise(workers = sum(S000))

# Make line chart
# AI-assisted
ggplot(lodes_county, aes(x = year, y = workers)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::comma, limits = c(0, NA)) +
  labs(
    title = "People who both work and live in Delaware County, 2013-2023",
    x = "Year",
    y = "Number of Workers"
  ) +
  theme_minimal()

```

```{r}
#| include: false

# AI-assisted

library(lehdr)

od_pa_main <- grab_lodes(
  state = "pa",
  year = 2023,
  lodes_type = "od",
  job_type = "JT00",
  segment = "S000",
  state_part = "main"
)

od_pa_aux <- grab_lodes(
  state = "pa",
  year = 2023,
  lodes_type = "od",
  job_type = "JT00",
  segment = "S000",
  state_part = "aux"
)

# Delaware county workers only
od_combined <- bind_rows(od_pa_main, od_pa_aux) |>
  filter(str_sub(h_geocode, 1, 5) == "42045") 

od_flow <- od_combined |>
  mutate(w_county = str_sub(w_geocode, 1, 5)) |>
  group_by(year, w_county) |>
  summarise(workers = sum(S000), .groups = "drop")

library(tidycensus)

# county name and county code
fips_lookup <- fips_codes |>
  mutate(fips = paste0(state_code, county_code)) |>
  select(fips, county, state)

# join od_flow
od_flow <- od_flow |>
  left_join(fips_lookup, by = c("w_county" = "fips"))

od_flow_23 <- od_flow |>
  arrange(desc(workers))

write_csv(od_flow_23, "../data/processed/od_flow_23.csv")
```


```{r}
#| label: Get LODES data in 2013
#| include: false
od_pa_main_13 <- grab_lodes(
  state = "pa",
  year = 2013,
  lodes_type = "od",
  job_type = "JT00",
  segment = "S000",
  state_part = "main"
)

od_pa_aux_13 <- grab_lodes(
  state = "pa",
  year = 2013,
  lodes_type = "od",
  job_type = "JT00",
  segment = "S000",
  state_part = "aux"
)

od_flow_13 <- bind_rows(od_pa_main_13, od_pa_aux_13) |>
  filter(str_sub(h_geocode, 1, 5) == "42045") |>
  mutate(w_county = str_sub(w_geocode, 1, 5)) |>
  group_by(year, w_county) |>
  summarise(workers = sum(S000), .groups = "drop") |>
  left_join(fips_lookup, by = c("w_county" = "fips")) |>
  arrange(desc(workers)) |>
  write_csv("../data/processed/od_flow_13.csv")
```

```{r}
#| label: Compare how commute number change in the past decade in Delaware County
#| echo: true
od_flow_change <- od_flow_13 |>
  select(w_county, county, state, workers_13 = workers) |>
  left_join(
    od_flow_23 |> select(w_county, workers_23 = workers),
    by = "w_county"
  ) |>
  mutate(
    change = workers_23 - workers_13,
    pct_change = (workers_23 - workers_13) / workers_13 * 100
  ) |>
  arrange(desc(change))

od_flow_change_table <- od_flow_change |>
  select(w_county, county, change, pct_change)

od_flow_change_table |> print (n =Inf)

write_csv(od_flow_change_table, "../data/processed/od_flow_change.csv")

```


```{r}
 del_all <- read.csv("../data/raw/lodes_del_any_23_13.csv") |> 
  clean_names()

glimpse(del_all)

# unique(del_all$year)
# unique(del_all$h_geocode) 

# Clean data
del_all_flow <- del_all |>
  mutate(h_county = str_sub(h_geocode, 1, 5)) |>
  mutate(w_county = str_sub(w_geocode, 1, 5)) |> 
  group_by(year, w_county) |>
  summarise(workers = sum(s000), .groups = "drop") |>
  left_join(fips_lookup, by = c("w_county" = "fips")) 

# Mutate change and percantage change number
del_all_flow_wide <- del_all_flow |>
  filter(year %in% c(2013, 2023)) |>
  pivot_wider(names_from = year, values_from = workers, names_prefix = "workers_") |>
  mutate(
    change = workers_2023 - workers_2013,
    pct_change = (workers_2023 - workers_2013) / workers_2013 * 100
  ) |>
  arrange(desc(change))

del_all_flow |> 
  print (n =Inf)

```

```{r}
# Total number of workers in 2023 and 2013
del_all_flow |>
  filter(year %in% c(2013, 2023)) |>
  group_by(year) |>
  summarise(total_workers = sum(workers))
```
```{r}
# share change
del_all_flow_wide <- del_all_flow_wide |>
  mutate(
    share_2013 = workers_2013 / sum(workers_2013, na.rm = TRUE),
    share_2023 = workers_2023 / sum(workers_2023, na.rm = TRUE) 
  ) |>
  mutate(share_change = share_2023 - share_2013)

```

```{r}
del_all_top_10 <- del_all_flow_wide |>
  slice_max(workers_2023, n = 10) |> 
  select(w_county, county, share_2013, share_2023, workers_2013, workers_2023, change) |> 
  write_csv("../data/processed/del_all_top_10.csv")

del_all_top_10_column <- del_all_flow_wide |>
  slice_max(workers_2023, n = 10) |> 
  select(county,state,workers_2013, workers_2023) |> 
  bind_rows(
    tibble(
      county = "Total",
      state = NA,
      workers_2013 = sum(del_all_flow_wide$workers_2013, na.rm = TRUE),
      workers_2023 = sum(del_all_flow_wide$workers_2023, na.rm = TRUE)
    )
  ) |>
  write_csv("../data/processed/del_all_top_10_column.csv")
```


::: {.callout-note collapse="false"}
## About the data
- BLS - Quarterly Census of Employment and Wages (QCEW)
- Source: https://data.bls.gov/cew/apps/table_maker/v4/table_maker.htm#type=7&year=2025&qtr=A&own=5&area=42045&supp=0
- NAICS Code instruction: https://www.bls.gov/iag/tgs/iag_index_naics.htm
- data dictionary: https://www.bls.gov/cew/additional-resources/open-data/csv-data-slices.htm 
- file download: http://www.bls.gov/cew/data/api/2025/a/area/42045.csv
:::



